Motional Ground State Cooling Outside the Lamb-Dicke Regime

### 2017-08-14 paper.tex
Kang-Kuen's edit and white space fix

%Systems of trapped neutral atoms assembled from the bottom-up in an array of optical tweezers are a promising  platform to study quantum information and quantum simulations~
Trapped neutral atoms, assembled  in an array of optical tweezers, are a promising  platform to study quantum information and quantum simulations~
\cite{Schlosser2001,Weiss2004,Isenhower2010,Wilk2010,Kaufman2015,Labuhn2016,Murmann2015}.
The inherent single-particle detection and control, combined with tunable interactions,
allow implementations of neutral-atom-based quantum logic gates~\cite{Isenhower2010,Wilk2010},
novel quantum phases~\cite{Labuhn2016}, and single-photon switches~\cite{Dayan2008,Tiecke2014}.
Advances in real-time re-arrangement of optical tweezers enable rapid preparation of atoms
in large and complex geometries~\cite{Barredo2016,Endres2016}.
Quantum motional control of individual
atoms~\cite{Li2012,Kaufman2012,Thompson2013,Liu2017,Robens2017} enable
studies of the atomic Hong-Ou-Mandel effect~\cite{Kaufman2014},
high-fidelity single qubit gates~\cite{Wang2016},
and efficient coupling of single atoms to photonic crystal cavities~\cite{Thompson2013a}.

@@

Preparation of single atoms in the motional ground state has been achieved in optical tweezers~\cite{Kaufman2012,Thompson2013,Liu2017,Robens2017}
 using Raman sideband cooling (RSC)~\cite{Monroe1995,Kerman2000,Han2000}.
However, RSC in these systems takes place in the Lamb-Dicke (LD) regime, where
the spread of the initial wavefunctions is  smaller than the reduced wavelength of light ($\lambdabar$)
used to address them.
Systems with a large RMS position spread ($z_{rms}$), either due to small mass or high initial temperature,
fall outside of the LD regime and undergo strong recoil heating during RSC.
In this letter we demonstrate cooling of single sodium atoms in optical tweezers,
initially outside  the LD regime ($z_{rms}/\lambdabar=4.0$), to the motional ground state.
We achieve a 3D ground state population of $P_0=81(4)$\% by  cooling via
 high-order sidebands in a carefully optimized cooling sequence.
Our approach is general and applicable to %opens up ground state cooling for
other systems.  Extending RSC beyond the LD regime opens  the possibility
for ground state cooling of systems such as light atoms and %systems with poor cooling such as
molecules that can be laser-cooled.

@@

During optical pumping, photon recoil can lead to significant heating, especially along the weakly-confined axial direction ($z$).  For an atom starting in an axial motional level $n_{\textrm{init}}$, the probability to end in another axial state $n$ after absorbing and re-emitting one photon is determined by an angular integral of the squared matrix element $|\langle n_{\textrm{init}}|e^{i \Delta k_{OP} z}| n\rangle|^2$ \cite{ItanoWineland1979}. Here $\Delta k_{OP}$ is the wave vector difference between absorbed and emitted photons, projected onto the $z$-axis.  As shown in Fig.~\ref{f-ld}A, the probability of gaining motional energy during OP grows with $n_{\textrm{init}}$, and is an important effect for sideband-cooling outside the Lamb-Dicke regime.

@@

Combining the axial and radial cooling results,
 a single Na atom is in  the 3D ground state  with a probability of 81(4)\% after RSC.
 We note that there is an additional 5\% population loss from  imaging  and 10\% loss
from  high-lying initial motional states that are not efficiently cooled by RSC,
but instead heated out of the trap by off-resonant scattering from the Raman beams.
The cooling efficiency is limited by spontaneous scattering rate (3-15 kHz) from the Raman beams, as well as spectral broadening from magnetic field fluctuations and trap anharmonicity.

>>>>

% Systems of trapped neutral atoms assembled from the bottom-up in an array of optical tweezers are a promising platform to study quantum information and quantum simulations~
Trapped neutral atoms, assembled in an array of optical tweezers, are a promising platform to study quantum information and quantum simulations~
\cite{Schlosser2001,Weiss2004,Isenhower2010,Wilk2010,Kaufman2015,Labuhn2016,Murmann2015}.
The inherent single-particle detection and control, combined with tunable interactions,
allow implementations of neutral-atom-based quantum logic gates~\cite{Isenhower2010,Wilk2010},
novel quantum phases~\cite{Labuhn2016}, and single-photon switches~\cite{Dayan2008,Tiecke2014}.
Advances in real-time re-arrangement of optical tweezers enable rapid preparation of atoms
in large and complex geometries~\cite{Barredo2016,Endres2016}.
Quantum motional control of individual
atoms~\cite{Li2012,Kaufman2012,Thompson2013,Liu2017,Robens2017} enable
studies of the atomic Hong-Ou-Mandel effect~\cite{Kaufman2014},
high-fidelity single qubit gates~\cite{Wang2016},
and efficient coupling of single atoms to photonic crystal cavities~\cite{Thompson2013a}.

@@

Preparation of single atoms in the motional ground state has been achieved in optical tweezers~\cite{Kaufman2012,Thompson2013,Liu2017,Robens2017}
 using Raman sideband cooling (RSC)~\cite{Monroe1995,Kerman2000,Han2000}.
However, RSC in these systems takes place in the Lamb-Dicke (LD) regime, where
the spread of the initial wavefunctions is smaller than the reduced wavelength of light ($\lambdabar$)
used to address them.
Systems with a large position spread ($z_{rms}$), either due to small mass or high initial temperature,
fall outside of the LD regime and undergo strong recoil heating during RSC.
In this letter we demonstrate cooling of single sodium atoms in optical tweezers,
initially outside the LD regime ($z_{rms}/\lambdabar=4.0$), to the motional ground state.
We achieve a 3D ground state population of $P_0=81(4)$\% by cooling via
 high-order sidebands in a carefully optimized cooling sequence.
Our approach is general and applicable to %opens up ground state cooling for
other systems. Extending RSC beyond the LD regime opens the possibility
for ground state cooling of systems such as light atoms and %systems with poor cooling such as
molecules that can be laser-cooled.

@@

During optical pumping, photon recoil can lead to significant heating, especially along the weakly-confined axial direction ($z$).  For an atom starting in an axial motional level $n_{\textrm{init}}$, the probability to end in another axial state $n$ after absorbing and re-emitting one photon is approximated by an angular integral of the squared matrix element $|\langle n_{\textrm{init}}|e^{i \Delta k_{OP} z}| n\rangle|^2$ \cite{ItanoWineland1979}. Here $\Delta k_{OP}$ is the wave vector difference between absorbed and emitted photons, projected onto the $z$-axis.  As shown in Fig.~\ref{f-ld}A, the probability of gaining motional energy during OP grows with $n_{\textrm{init}}$, and is an important effect for sideband-cooling outside the Lamb-Dicke regime.

@@

Combining the axial and radial cooling results,
 a single Na atom is in the 3D ground state with a probability of 81(4)\% after RSC.
 We note that there is an additional 5\% population loss from imaging and 10\% loss
from high-lying initial motional states that are not efficiently cooled by RSC,
but instead heated out of the trap by off-resonant scattering from the Raman beams.
The cooling efficiency is limited by spontaneous scattering rate (0.5-1 kHz) from the Raman beams, as well as spectral broadening from magnetic field fluctuations and trap anharmonicity.
